---
title: "Preventable deaths - Focus on suicide in the United States, 2017"
author: "[Laurel  Boyd](https://github.com/laurelboyd)"
date: "`r Sys.Date()`"
output: xaringan::moon_reader
#yolo: true #under 
---

```{r setup, include=FALSE, messages = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```
```{r install required packages}
#devtools::install_github("yihui/xaringan")
library(xaringan) #automatically update markdown when editing
library(arules) #create sparse matrix
#library(dplyr) #SQL wrapper
#library(stringr) #regex functionality within dplyr
```

## Suicide in the United States

According to the CDC, suicide is the 10th leading cause of death, making it one of the most preventable deaths in the United States.[^1] This investigation will look at causes of death associated with suicide to investigate where intervention (purely on the basis of combinations of comborbities) could possibly be applied. 

[^1] https://www.cdc.gov/violenceprevention/suicide/fastfact.html

Causes of death for 2017 (most recent year of available data) have been imported directly from the CDC Wonder website: https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm#Mortality_Multiple . They have been imported via a python script (to account for file format without delimeters) and the processed data is imported below.

With just causes of death columns, subset of data we can look at combinations of cause of death.

First we need to import the CSV file with causes of death - but to consolidate the causes of death into a single column (create a sparse matrix by removing the commas)
```{r import data}
VS17 <-read.transactions("VS17MORT.csv", sep= ",")
```
Next we can look at the overal number and distrubution of causes of death.

```{r overall death summary}
summary(VS17)
```
There were 5,640,070 deaths reported in 2017 using 5661 ICD-10 codes. Cardiac arrest (ICD-10 I46.9) is the most commonly reported cause of death followed by hypertention (I10).

causes of suicide

Next, we'll subset all reported deaths to those of interest for this investigation, ICD-10 codes associated with suicide: U03,X60-X84, and Y87.0.

```{r subset to suicide codes}
# 
# diabetesrules <-subset(VS17, items %pin% c("E11"))
# summary(diabetesrules)
# 
# 
# eatingdisorderrules <-subset(VS17, items %pin% c("F509") | items %pin% c("E43"))
# summary(eatingdisorderrules)
# 
# tobaccorules <-subset(VS17, items %pin% c("F179"))
# summary(tobaccorules)
# 
# 
# injuryrules <-subset(VS17, items %pin% c("T149") | items %pin% c("TS099")| items %pin% c("T09"))
# summary(injuryrules)

depressionrules <-subset(VS17, items %pin% c("F32"))
summary(depressionrules)

# nicotinerules <-subset(VS17, items %pin% c("F17"))
# summary(nicotinerules)
# suiciderules <-subset(VS17, items %pin% c("U03") |items %pin% c("X6") | items %pin% c("X7") | items %pin% c("X81") | items %pin% c("X82")
#                         | items %pin% c("X83") | items %pin% c("X84") | items %pin% c("Y87"))
# 
# summary(suiciderules)
```

There were 33,524 deaths with depression listed as a cause of death in 2017. Of these, most common codes associated with the depression diagnosis are I10 (essential hypertension), F03 (unspecified dementia) and F41.9 (unspecified anxiety disorder).

```{r frequency plots}
par(mfrow=c(1,2))
options("scipen"=1000, "digits"=6)
itemFrequencyPlot(depressionrules,
                  type="relative",
                  topN=10, # can be changed to the number of interest
                  horiz=TRUE,
                  col='steelblue3',
                  xlab='',
                  main='Item frequency, relative')

itemFrequencyPlot(depressionrules,
                  type="absolute",
                  topN=10,
                  horiz=TRUE,
                  col='steelblue3',
                  xlab='',
                  main='Item frequency, absolute')


# tbl <- crossTable(depressionrules, sort=TRUE)
# tbl[1:5,1:5]
```
Next we'll look at lift of combinations of diagnoses associated with nicotine dependence. Broadly speaking, lift greater than 1 is indicative of complementary combinations while lift less than 1 is indicative of substitute combinations.

```{r crosstabs}
crossTable(depressionrules, measure='lift',sort=T)[1:10,1:10]
```

Run a chi-square test to see whether the association between depression and nicotine dependence is statistically significant (as lift is close to 1 but the overall counts are around 5,000 so quite high).

```{r search for rules, echo=FALSE}
crossTable(depressionrules, measure='chi')['F329', 'F179']
```
It's likely the probability of nicotine dependence assigned if depression is assigned is not due to chance alone. The chi-square test is statistically signficant.

This is an interesting insight but it's likely there are other combinations of rules.

```{r look for rules}

rules <- apriori(depressionrules,
                 parameter = list(support=.00005,
                                  minlen=2,
                                  target='rules' # to mine for rules
                                  ))
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
